Bayesian Sparsification Methods for Deep Complex-valued Networks

### 2020-06-23 paper/paper.tex
proofread the paper

% applications of cvnn
Complex-valued nature of the data in acoustic and radio signal processing has been
the main driver behind the adoption of $\cplx$-valued neural networks ($\cplx$VNN).
\citet{hirose_complex-valued_2009} argues that the combined phase-magnitude effect of
$\cplx$-valued transformations reduces removes the excess degrees of freedom, that cause
degenerate transformation in $\real$-valued networks with twice the feature dimensions.
Their study demonstrates superiority of $\cplx$VNN in landmine detection using ground
penetrating radar imaging. Other examples, where $\cplx$-valued networks have outperformed
$\real$-valued networks include magnetic resonance \citep{hui_mri_1995,wang_deepcomplexmri_2020}
and radar imaging \citep{haensch_complex-valued_2010,zhang_complex-valued_2017}, music
transcription and spectral speech modelling \citep{wisdom_full-capacity_2016,trabelsi_deep_2018},
and wireless signal classification \citep{yang_complex_2019}. \citet{tarver_design_2019}
have lowered the out-of-band power leakage with a $\cplx$-valued network for digital
signal predistortion. The networks have also been applied to non-$\cplx$-valued domains,
such as image classification \citep{popa_complex-valued_2017}, sequence modelling
\citep{danihelka_associative_2016}, and motion prediction \citep{wolter_complex_2018},
and for stabilizing back-propagation in RNN \citep{wisdom_full-capacity_2016}.

% motivation and other methods
Despite promising results for embedded signal processing applications, $\cplx$-valued
networks remain a niche in deep learning, and as such little attention has been paid
to compression methods specific to $\cplx$VNN. \citet{wu_compressing_2019} adapt $k$-means
quantization to $\cplx \simeq \real^2$ parameters and successfully compress $\cplx$VNN
with the ``prune-quantize-code'' procedure of \citet{han_deep_2016}.

Yet there is an abundance of research related to real-valued network compression, and
many results can be applied to $\cplx$VNN. Methods such as knowledge distillation
\citep{hinton_distilling_2015}, which trains a small network to replicate a large
well-trained teacher, low-rank matrix \citep{denton_exploiting_2014} and tensor decomposition
\citep{novikov_tensorizing_2015}, or magnitude-based parameter pruning \citep{zhu_prune_2018}
can be adapted to $\cplx$VNN without modifications. Parameter quantization and conversion
from floating to fixed point arithmetic \citep{courbariaux_training_2015,uhlich_mixed_2020},
appear to be readily applicable as well. \citet{louizos_learning_2018} propose probabilistic
$\ell_0$ regularization through multiplicative $[0, 1]$-valued stochastic masks with
distribution having an atom at $0$, yet differentiable via the reparameterization trick
\citep{kingma_auto-encoding_2014}. By sharing a single mask value within a group of
parameters their approach can be adapted to $\cplx$ parameters. However, methods such
as Hessian-based parameter pruning \citep{lecun_optimal_1990} or Sparse Variational Dropout
\citep{molchanov_variational_2017} require additional considerations.
% the more explicitly zero parameters there is, the less computations is needed.

@@

The paper is structured as follows. Sec.~\ref{sec:variational_dropout} reviews Variational
Dropout, and sec.~\ref{sec:c_valued_networks} provides a brief summary of the inner
working of complex-valued networks. The main contribution of this study is presented in
sec.~\ref{sec:c_variational_dropout}, where we outline the details of different variational
approximations and priors and derive the penalty terms in the evidence lower bound.
In sec.~\ref{sec:experiments} we evaluate the sparsification rate, compare the resulting
performance of various $\cplx$-networks proposed in prior work, and discuss the outcomes.

@@

For an observed dataset $
  D = (x_i)_{i=1}^N
$ and statistical model $
  p(D \mid \omega)
    = \prod_i \log p(x_i \mid \omega)
    % $D$ conditional independence given parameters $\omega$
$ with parameters $\omega$ the Bayes rule transforms prior hypotheses $\pi(\omega)$
about the unknown distribution of model's parameters into the posterior distribution: $
  p(\omega \mid D) = \tfrac{p(D \mid \omega) \pi(\omega)}{p(D)}
$.
%
% In general the marginal likelihood of the dataset is intractable $
%   p(D) = \mathbb{E}_{\pi(\omega)} p(D \mid \omega)
% $.
Save for the relatively simple set-ups, either the posterior distribution itself or the
mathematical expectations it is involved in are analytically intractable or impractical
to compute numerically. Variational Inference (VI), proposed by \citet{jordan_introduction_1999},
can be used in such cases to make approximate inference. The approach finds an approximation
within some distribution family $q_\theta(\omega)$, which is closest to the true posterior
distribution in terms of Kullback-Leibler divergence: $
  KL(q_\theta(\omega) \| p(\omega \mid D))
    % = \int \frac{dQ}{dP} \log{\frac{dQ}{dP}} P(d\omega)
    % = \int \frac{q(\omega)}{p(\omega)} \log{\frac{q(\omega)}{p(\omega)}} p(\omega) d\omega
    % = \mathbb{E}_{\omega \sim q}
    %   \log\frac{q(\omega)}{p(\omega)}
    = \mathbb{E}_{\omega \sim q_\theta}
      \log \tfrac{q_\theta(\omega)}{p(\omega \mid D)}
$. % The approximation family is chosen so that the expectations become tractable.
%
\citet{jordan_introduction_1999} show that this problem is equivalent to variational
\emph{maximization} of the \emph{Evidence Lower Bound} (ELBO)
\begin{equation}  \label{eq:elbo_general}
  \mathcal{L}(\theta; \lambda)
    = - KL(q_{\theta} \| \pi_{\lambda})
      + \sum_{i=1}^N \mathbb{E}_{\omega \sim q_{\theta}}
        \log p(x_i \mid \omega)
      % \mathbb{E}_{\omega \sim q_{\theta}}
        % \log\frac{q_\theta(\omega)}{\pi_{\lambda}(\omega)}
  % = \mathbb{E}_{\omega \sim q_{\theta}}
  %     \log{p_{\phi}(D \mid \omega) \pi_{\lambda}(\omega)}
  %   + \mathbb{H}(q_{\theta})
  \,,
\end{equation}
where the variational parameters $\theta$ and $\lambda$ parameterize the approximation
and the prior, respectively.
%
Kullback-Leibler and, by proxy, ELBO are standard objectives in VI, however it is possible
to use other objectives, provided the true posterior $p(\omega \mid D)$ is evaluated only
through $\log p(D \mid \omega)$ and $\log \pi(\omega)$, \citep{ranganath_operator_2016}.

@@

% The Local Reparameterization Trick reduces the variance of SGVB
% even further by sampling independent weight matrices for
% each data-point inside mini-batch. So we need both original SGVB
% (per sample randomness) and local reparameterization (that is an
% efficient and less noisy equivalent to per-sample weight sampling)
% SGVB is specifically for VAE (local latent variables per data-point)
% what Kingma propose is SGVB for global shared latent variables aka
% WEIGHTS!!
In this study we use the SGVB estimator \eqref{eq:sgvb_estimator} with $L=1$ and the
\emph{local reparameterization trick} proposed by \citet{kingma_variational_2015}. They
argued that this gradient estimator can be made more statistically and computationally
efficient, if the structure of the model permits translating global stochasticity of
$\omega$ down to local intermediate states of computation.
%
The class of models that allow this include non-recurrent computational graphs,
exemplified by neural networks with parameters $\omega \sim q_\theta$. In this case
\eqref{eq:sgvb_estimator} would require that the \emph{entire} set of network's parameters
$\omega$ be independently drawn for each element in the mini-batch. However, noting
that the parameters in a network naturally decompose into subsets with non-overlapping
layer-wise effects, it is standard to assume that the approximation $
  q_\theta(\omega)
$ is factorized over layers. Furthermore, if $
  % \omega =
  W \in \mathbb{R}^{n\times m}
$ in a linear layer $
  y = b + W^\top x
$ with $
  q_\theta(W)
    = \prod_{ij}
      \mathcal{N}(w_{ij} \vert\, \mu_{ij}, \sigma_{ij}^2)
$, then by virtue of $y$ being a linear transformation of $W$, we get
\begin{equation}  \label{eq:r-gauss-trick}
  q(y) = \prod_i \mathcal{N}\Bigl(
        y_i \big\vert\,
        b_i + \sum_j \mu_{ij} x_j,
        \sum_j \sigma^2_{ij} x_j^2
    \Bigr)
  \,.
\end{equation}
This yields outputs equivalent in distribution to sampling $W$ for each element in
the mini-batch, which produces the SGVB estimator with smaller variance, as demonstrated
by \citet{kingma_variational_2015}.
% Similar idea to \eqref{eq:r-gauss-trick} was proposed by \citet{wang_fast_2013} to speed
% up Gaussian dropout.

@@

Due factorization assumption, the term $KL(q_\theta \| \pi)$ in \eqref{eq:sgvb_estimator}
for Sparse VD unravels into $
  \sum_j K(\tfrac{\sigma^2_{j}}{\mu_{j}^2})
$ with
\begin{equation}  \label{eq:improper-kl-div-real}
  K(\alpha)
    \propto \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
        \log{\Bigl\lvert \tfrac1{\sqrt{\alpha}} + \varepsilon \Bigr\rvert^2}
      % \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
      %   \log{\bigl\lvert 1 + \sqrt{\alpha} \varepsilon \bigr\rvert^2}
      % - \frac12 \log{\alpha}
  \,.
\end{equation}
\citet{kingma_variational_2015} approximated $K(\alpha)$ over $\alpha \in (0, 1)$ by
a polynomial with $\log$-term, and later \citet{molchanov_variational_2017} refined the
approximation of \eqref{eq:improper-kl-div-real} by weighted sum of a sigmoid and a
soft-plus term.
% see p.6 sec 3.4 of kingma_variational_2015
In appendix~\ref{sec:real-chisq-grad} we verify the value and the derivative of their
approximation against a Monte-Carlo estimate for
$\alpha$ varying over a fine logarithmic grid and the gradient of \eqref{eq:improper-kl-div-real},
for which we obtain an exact expression.
% (!) good for SGVB: gradient-based methods require unbiased gradient estimates
% (in) it is in poor taste to brag about this beign the first time anyone has derived a
% derivative of an intractable function.

@@

% implementation
$\cplx$-valued neural networks are networks that rely on the arithmetic in the complex
domain. To achieve this implementations of $\cplx$VNN use a geometric representation
of a complex number as paired \emph{real} and \emph{imaginary} values, $\cplx \simeq \real^2$,
ensuring that the resulting $\real$-valued computational graph respects $\cplx$-arithmetic.
For example, $
  f\colon \cplx^n \to \cplx^m
$ is identified with a real vector-valued function $
  F\colon \real^{2 m} \to \real^{2 m}
$ defined via $
  F(u, v) = (\Re f(u + \iu v), \Im f(u + \iu v))
$, $\Re$ and $\Im$ denoting the real and imaginary parts, respectively. When $f$ is
a $\cplx$-valued linear transformation, the computations are ``wired'' so that
\begin{equation}  \label{eq:cplx-lin-op}
  % F\colon
  %   (u, v)
  %     \mapsto \Bigl(
  %       P u - Q v,
  %       P v + Q u
  %     \Bigr)
  F(u, v)
    =
    \begin{pmatrix}
      P u - Q v \\
      P v + Q u
    \end{pmatrix}
    = \begin{pmatrix}
      P & - Q \\ Q & P
    \end{pmatrix} \begin{pmatrix}
      u \\ v
    \end{pmatrix}
    \,,
\end{equation}
with $
  P, Q \colon \real^{n} \to \real^{m}
$ given by $P = \Re f$ and $Q = \Im f$ restricted to $\real^{n}$.
% The downside is that, for instance, one $\cplx$-valued convolution requires up to
% four $\real$-valued convolution operations.
%
Non-linearities in $\cplx$VNN can be hyperbolic functions or maps that operate on
$\cplx$ numbers in planar form, $
  z \mapsto \sigma(\Re{z}) + \iu \sigma(\Im{z}) % split
$, or polar form $
  % z \mapsto \sigma(\lvert z\rvert, \arg{\!(z)})
  r e^{\iu \phi} \mapsto \sigma(r, \phi)
$.
% \citep{savitha_2011,hirose_generalization_2012,arjovsky_unitary_2016,guberman_complex_2016}

This $\cplx \simeq \real^2$ identification allows straightforward retrofitting of $\cplx$VNN
into existing $\real$-valued auto-differentiation frameworks for deep learning. This
act is backed by Wirtinger ($\cplx\real$) calculus, which enables generalized treatment
of functions of complex argument, by regarding $z$ and its \emph{complex conjugate}
$\conj{z}$ as independent variables and defining derivatives with respect to them through
partial derivatives with respect to real and imaginary parts. These definitions simplify
manual calculations of $\cplx$ derivatives and satisfy the product and chain rules, respect
complex conjugation and linearity for $\cplx \to \cplx$ maps. $\cplx\real$ calculus was
used to define $\cplx$ version of back-propagation,
\citep{benvenuto_complex_1992,guberman_complex_2016}.
However, since auto-differentiation frameworks can algorithmically handle computational
graphs of arbitrary complexity, explicit use of Wirtinger derivatives is not required,
especially considering the fact that the direction of the steepest ascent of a $
  \cplx \to \real
$ function is given by complex conjugate gradient $\nabla_{\conj{z}}$, which coincides
with the classical gradient of the same function viewed as $\real^2 \to \real$,
(see appendix~\ref{sub:wirtinger_calculus}).
%
% in natively $\cplx$-valued backprop it would be necessary to modify the existing sgd
% to explicitly use $\nabla_{\conj{z}} f(z)$, while no such modification is necessary
% in re-im. If some inner activation of a $\cplx$VNN is holomorphic, i.e. satisfies
% Cauchy-Riemann conditions, then that particular step of in the chain rule that is
% backprop can be simplified by assuming one piece of the tracked gradient to be exact
% zero. But other wise it is necessary to track both $\partial_{\conj{z}}$ and $\partial_{z}$.
%
% Training a $\cplx$VNN involves optimizing a real-valued objective with respect to
% complex valued arguments.
%
% The computed gradients are theoretically justified
% - CR dictates the it is necessary to track both $z$ and $\conj{z}$ derivatives
% - the direction of largest increase in f: C-R is given by the gradient with
% respect to the conjugate, which coincides in the re-im representation with 
% the ordinary gradient wrt u and v
% - the Chain Rule in CR holds and when a holomorphic C-C function is backproped
% the autodiff just wastes cycles on computing its conj-derivative (rather than assuming
% it to be zero)

% Although $\cplx$-valued dense layers and back-propagation have been around since early
% 1990-s, one of the earliest accounts
Development of deep $\cplx$-valued networks has been active. \citet{haensch_complex-valued_2010}
put forward $\cplx$-valued convolutional networks, \citet{guberman_complex_2016} and
\citet{popa_complex-valued_2017} developed modifications of pooling, \citet{arjovsky_unitary_2016}
and \citet{wisdom_full-capacity_2016} proposed $\cplx$-valued RNNs with unitary recurrent
transition matrices, and \citet{danihelka_associative_2016} developed $\cplx$-valued
holographic representations for LSTMs. More recently \citet{trabelsi_deep_2018} proposed
$\cplx$-valued batch-normalization and weight initialization, \citet{wolter_complex_2018}
investigated different $\cplx$-valued gating mechanisms for RNNs, \citet{yang_complex_2019}
proposed $\cplx$-valued self-attention and complex transformer architecture. It merits
noting that \citet{gaudet_deep_2018} generalized $\cplx$VNN further to deep quaternion-valued
networks, and \citet{vecchi_compressing_2020} studied sparsity inducing regularizers for
them.
% complex-valued distributions \citep{pav_moments_2015,taubock_complex-valued_2012},
% and \citep{karseras_caution:_2014}

@@

Below we develop Sparse Variational Dropout for $\cplx$VNN by using fully factorized
complex Gaussian posterior approximation, outline the $\cplx$ version of the local
reparameterization trick and derive the divergence penalties in \eqref{eq:sgvb_estimator}.
The proposed $\cplx$-valued extension can readily be a part of a hierarchical variational
approximation for structured sparsity \citep{louizos_bayesian_2017}.

@@

$, respectively, and $z$ is a \emph{circularly symmetric} $\cplx$-Gaussian vector
if $z$ and $\conj{z}$ are uncorrelated, i.e. $C = 0$.

@@

For a fully factorized approximation $q(\omega)$ and factorized prior $
  \pi(\omega) = \prod_{ij} \pi(\omega_{ij})
$, the divergence term in ELBO \eqref{eq:sgvb_estimator} is
\begin{equation}  \label{eq:elbo-general-kl-div}
  KL(q \| \pi)
    % = \mathbb{E}_{\oemga \sim q(\oemga)} \log \tfrac{q(\oemga)}{\pi(\oemga)}
    % = \sum_{ij} \mathbb{E}_{\omega_{ij} \sim q(\omega_{ij})}
    %   \log \tfrac{q(\omega_{ij})}{\pi(\omega_{ij})}
    % = \sum_{ij} KL\bigl( q(\omega_{ij}) \| \pi(\omega_{ij}) \bigr)
    = - \sum_{ij}
        \mathbb{H}(q(\omega_{ij}))
        + \mathbb{E}_{q(\omega_{ij})} \log{\pi(\omega_{ij})}
    \,.
\end{equation}
We consider two fully factorized priors: an improper prior resembling VD and $\cplx$-Gaussian
ARD prior. We omit subscripts ${ij}$ for brevity in this section.

@@

% write a note on the incompressible parameters (bias, batch norm etc.)
The compression rate is calculated based on the number of floating point values needed
to store the network and equals $
  \tfrac{n_\mathtt{par}}{n_\mathtt{par} - n_\mathtt{zer}}
$, where $n_\mathtt{zer}$ is the number of explicit zeros at the ``fine-tune'' stage
and $n_\mathtt{par}$ is the total number of values. In a $\real$-valued network each
parameter counts as \emph{one} value and as \emph{two} values in a $\cplx$VNN. Each
model has a compression limit, determined by biases, shift and scaling in $\real$-
and $\cplx$-valued batch normalization layers.

>>>>

% applications of cvnn
Complex-valued nature of the data in acoustic and radio signal processing has been
the main driver behind the adoption of $\cplx$-valued neural networks ($\cplx$VNN).
\citet{hirose_complex-valued_2009} argues that the combined phase-magnitude effect of
$\cplx$-valued transformations removes the excess degrees of freedom, that cause
degenerate transformations in $\real$-valued networks with twice the feature dimensions.
Their study demonstrates superiority of $\cplx$VNN in landmine detection using ground
penetrating radar imaging. Other examples, where $\cplx$-valued networks have outperformed
$\real$-valued networks, include magnetic resonance \citep{hui_mri_1995,wang_deepcomplexmri_2020}
and radar imaging \citep{haensch_complex-valued_2010,zhang_complex-valued_2017}, music
transcription and spectral speech modelling \citep{wisdom_full-capacity_2016,trabelsi_deep_2018},
and wireless signal classification \citep{yang_complex_2019}. \citet{tarver_design_2019}
have lowered the out-of-band power leakage with a $\cplx$-valued network for digital
signal predistortion. The networks have also been applied to non-$\cplx$-valued domains,
such as image classification \citep{popa_complex-valued_2017}, sequence modelling
\citep{danihelka_associative_2016}, and motion prediction \citep{wolter_complex_2018},
and for stabilizing back-propagation in RNN \citep{wisdom_full-capacity_2016}.

% motivation and other methods
Despite promising results for embedded signal processing applications, $\cplx$-valued
networks remain a niche in deep learning, and as such little attention has been paid
to compression methods specific to $\cplx$VNN.
%
Yet there is an abundance of research related to real-valued network compression, and
many results can be applied to $\cplx$VNN. Methods such as knowledge distillation
\citep{hinton_distilling_2015}, which trains a small network to replicate a large
well-trained teacher, low-rank matrix \citep{denton_exploiting_2014} and tensor decomposition
\citep{novikov_tensorizing_2015}, or magnitude-based parameter pruning \citep{zhu_prune_2018}
can be adapted to $\cplx$VNN without modifications. Parameter quantization and conversion
from floating to fixed point arithmetic \citep{courbariaux_training_2015,uhlich_mixed_2020},
appear to be readily applicable as well. For example, \citet{wu_compressing_2019} adapt
$k$-means quantization to $\cplx \simeq \real^2$ parameters and successfully compress
$\cplx$VNN with the ``prune-quantize-code'' procedure of \citet{han_deep_2016}.

Other methods cannot be translated to $\cplx$VNN this straightforwardly. Probabilistic
$\ell_0$ regularization of \citet{louizos_learning_2018} prune networks using multiplicative
$[0, 1]$-valued stochastic masks with distributions having an atom at $0$, yet differentiable
via the reparameterization trick \citep{kingma_auto-encoding_2014}. By sharing a single mask
value within a group of parameters their approach can be adapted to $\cplx$ parameters.
However, methods such as Hessian-based parameter pruning \citep{lecun_optimal_1990}
or Sparse Variational Dropout \citep{molchanov_variational_2017} require additional
considerations.
% the more explicitly zero parameters there is, the less computations is needed.

@@

The paper is structured as follows. Sec.~\ref{sec:variational_dropout} reviews Variational
Dropout, and sec.~\ref{sec:c_valued_networks} provides a brief summary of the inner
workings of complex-valued networks. The main contribution of this study is presented
in sec.~\ref{sec:c_variational_dropout}, where we provide the details of $\cplx$-valued
variational sparsification method. In sec.~\ref{sec:experiments} we estimate the compression
and performance trade-off on shallow and deep $\cplx$-valued networks, and discuss the
outcomes.

@@

For an observed dataset $
  D = (x_i)_{i=1}^N
$ and statistical model $
  p(D \mid \omega)
    = \prod_i p(x_i \mid \omega)
    % $D$ conditional independence given parameters $\omega$
$ with parameters $\omega$ the Bayes rule transforms prior hypotheses $\pi(\omega)$
about the unknown distribution of model's parameters into the posterior distribution: $
  p(\omega \mid D) = \tfrac{p(D \mid \omega) \pi(\omega)}{p(D)}
$.
%
% In general the marginal likelihood of the dataset is intractable $
%   p(D) = \mathbb{E}_{\pi(\omega)} p(D \mid \omega)
% $.
Save for the relatively simple set-ups, either the posterior distribution itself or the
mathematical expectations it is involved in are analytically intractable or impractical
to compute numerically. Variational Inference (VI), proposed by \citet{jordan_introduction_1999},
can be used in such cases to make approximate inference. The approach finds an approximation
within some distribution family $q_\theta(\omega)$, which is closest to the true posterior
distribution in terms of Kullback-Leibler divergence: $
  KL(q_\theta(\omega) \| p(\omega \mid D))
    % = \int \frac{dQ}{dP} \log{\frac{dQ}{dP}} P(d\omega)
    % = \int \frac{q(\omega)}{p(\omega)} \log{\frac{q(\omega)}{p(\omega)}} p(\omega) d\omega
    % = \mathbb{E}_{\omega \sim q}
    %   \log\frac{q(\omega)}{p(\omega)}
    = \mathbb{E}_{\omega \sim q_\theta}
      \log \tfrac{q_\theta(\omega)}{p(\omega \mid D)}
$. % The approximation family is chosen so that the expectations become tractable.
%
\citet{jordan_introduction_1999} show that this problem is equivalent to variational
\emph{maximization} of the \emph{Evidence Lower Bound} (ELBO)
\begin{equation}  \label{eq:elbo_general}
  \mathcal{L}(\theta; \lambda)
    = - KL(q_{\theta} \| \pi_{\lambda})
      + \sum_{i=1}^N \mathbb{E}_{\omega \sim q_{\theta}}
        \log p(x_i \mid \omega)
      % \mathbb{E}_{\omega \sim q_{\theta}}
        % \log\frac{q_\theta(\omega)}{\pi_{\lambda}(\omega)}
  % = \mathbb{E}_{\omega \sim q_{\theta}}
  %     \log{p_{\phi}(D \mid \omega) \pi_{\lambda}(\omega)}
  %   + \mathbb{H}(q_{\theta})
  \,,
\end{equation}
where the variational parameters $\theta$ and $\lambda$ parameterize the approximation
and the prior, respectively.
%
Kullback-Leibler and, by proxy, ELBO are standard objectives in VI, however it is possible
to use other objectives, provided the true posterior $p(\omega \mid D)$ is evaluated only
through $\log p(D \mid \omega)$ and $\log \pi(\omega)$, \citep{ranganath_operator_2016}.

@@

% The Local Reparameterization Trick reduces the variance of SGVB
% even further by sampling independent weight matrices for
% each data-point inside mini-batch. So we need both original SGVB
% (per sample randomness) and local reparameterization (that is an
% efficient and less noisy equivalent to per-sample weight sampling)
% SGVB is specifically for VAE (local latent variables per data-point)
% what Kingma propose is SGVB for global shared latent variables aka
% WEIGHTS!!
In this study we use the SGVB estimator \eqref{eq:sgvb_estimator} with $L=1$ and the
\emph{local reparameterization trick} proposed by \citet{kingma_variational_2015}. They
argued that this gradient estimator can be made more statistically and computationally
efficient, if the structure of the model permits translating global stochasticity of
$\omega$ down to local intermediate states of computation.
%
The class of models that allow this include non-recurrent computational graphs,
exemplified by neural networks with parameters $\omega \sim q_\theta$. In their case,
\eqref{eq:sgvb_estimator} would require that the \emph{entire} set of network's parameters
$\omega$ be independently drawn for each element in the mini-batch. Since that the
parameters in a network naturally split into subsets with non-overlapping layer-wise
effects, it is standard to assume that the approximation $
  q_\theta(\omega)
$ is factorized over layers. Furthermore, if $
  % \omega =
  W \in \mathbb{R}^{n\times m}
$ in a linear layer $
  y = b + W^\top x
$ with $
  q_\theta(W)
    = \prod_{ij}
      \mathcal{N}(w_{ij} \vert\, \mu_{ij}, \sigma_{ij}^2)
$, then by virtue of $y$ being a linear transformation of $W$, we get
\begin{equation}  \label{eq:r-gauss-trick}
  q(y) = \prod_i \mathcal{N}\Bigl(
        y_i \big\vert\,
        b_i + \sum_j \mu_{ij} x_j,
        \sum_j \sigma^2_{ij} x_j^2
    \Bigr)
  \,.
\end{equation}
This yields outputs equivalent in distribution to sampling $W$ for each element in
the mini-batch, which produces the SGVB estimator with smaller variance, as demonstrated
by \citet{kingma_variational_2015}.
% Similar idea to \eqref{eq:r-gauss-trick} was proposed by \citet{wang_fast_2013} to speed
% up Gaussian dropout.

@@

Due factorization assumption, the term $KL(q_\theta \| \pi)$ in \eqref{eq:sgvb_estimator}
for Sparse VD unravels into $
  \sum_j K(\tfrac{\sigma^2_{j}}{\mu_{j}^2})
$ with
\begin{equation}  \label{eq:improper-kl-div-real}
  K(\alpha)
    \propto \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
        \log{\Bigl\lvert \tfrac1{\sqrt{\alpha}} + \varepsilon \Bigr\rvert^2}
      % \frac12 \mathbb{E}_{\varepsilon \sim \mathcal{N}(0, 1)}
      %   \log{\bigl\lvert 1 + \sqrt{\alpha} \varepsilon \bigr\rvert^2}
      % - \frac12 \log{\alpha}
  \,.
\end{equation}
\citet{kingma_variational_2015} approximated $K(\alpha)$ over $\alpha \in (0, 1)$ by a
polynomial with a logarithmic term, and later \citet{molchanov_variational_2017} refined
the approximation of \eqref{eq:improper-kl-div-real} by weighted sum of a sigmoid and a
soft-plus term.
% see p.6 sec 3.4 of kingma_variational_2015
In appendix~\ref{sec:real-chisq-grad} we verify the value and the derivative of their
approximation against a Monte-Carlo estimate for $\alpha$ varying over a fine logarithmic
grid and the exact expression for gradient of \eqref{eq:improper-kl-div-real}.
% (!) good for SGVB: gradient-based methods require unbiased gradient estimates
% (in) it is in poor taste to brag about this beign the first time anyone has derived a
% derivative of an intractable function.

@@

% implementation
$\cplx$-valued neural networks are networks that rely on the arithmetic in the complex
domain. To achieve this implementations of $\cplx$VNN use the geometric representation
of a complex number as paired \emph{real} and \emph{imaginary} values, $\cplx \simeq \real^2$,
ensuring that the resulting $\real$-valued computational graph respects $\cplx$-arithmetic.
For example, $
  f\colon \cplx^n \to \cplx^m
$ is identified with a real vector-valued function $
  F\colon \real^{2 m} \to \real^{2 m}
$ defined via $
  F(u, v) = (\Re f(u + \iu v), \Im f(u + \iu v))
$, $\Re$ and $\Im$ denoting the real and imaginary parts, respectively. When $f$ is
a $\cplx$-valued linear transformation, the computations are ``wired'' so that
\begin{equation}  \label{eq:cplx-lin-op}
  % F\colon
  %   (u, v)
  %     \mapsto \Bigl(
  %       P u - Q v,
  %       P v + Q u
  %     \Bigr)
  F(u, v)
    =
    \begin{pmatrix}
      P u - Q v \\
      P v + Q u
    \end{pmatrix}
    = \begin{pmatrix}
      P & - Q \\ Q & P
    \end{pmatrix} \begin{pmatrix}
      u \\ v
    \end{pmatrix}
    \,,
\end{equation}
with $
  P, Q \colon \real^{n} \to \real^{m}
$ given by $P = \Re f$ and $Q = \Im f$ restricted to $\real^{n}$.
% The downside is that, for instance, one $\cplx$-valued convolution requires up to
% four $\real$-valued convolution operations.
%
Non-linearities in $\cplx$VNN can be hyperbolic functions or maps that operate on
$\cplx$ numbers in planar form, $
  z \mapsto \sigma(\Re{z}) + \iu \sigma(\Im{z}) % split
$, or polar form $
  % z \mapsto \sigma(\lvert z\rvert, \arg{\!(z)})
  r e^{\iu \phi} \mapsto \sigma(r, \phi)
$.
% \citep{savitha_2011,hirose_generalization_2012,arjovsky_unitary_2016,guberman_complex_2016}

This $\cplx \simeq \real^2$ identification allows straightforward retrofitting of $\cplx$VNN
into existing $\real$-valued auto-differentiation frameworks for deep learning. This
act is backed by Wirtinger ($\cplx\real$) calculus, which enables generalized treatment
of functions of complex argument, by regarding $z$ and its \emph{complex conjugate}
$\conj{z}$ as independent variables and defining derivative operators with respect to
them through partial derivatives with respect to real and imaginary parts. These definitions
simplify manual analysis of $\cplx$ derivatives and satisfy the product and chain rules, respect
complex conjugation and linearity for $\cplx \to \cplx$ maps, and such were used to define
$\cplx$ version of back-propagation, \citep{benvenuto_complex_1992,guberman_complex_2016}.
However, since auto-differentiation frameworks can algorithmically handle computational
graphs of arbitrary complexity, explicit use of Wirtinger derivatives is not required,
especially considering the fact that the direction of the steepest ascent of a $
  \cplx \to \real
$ function is given by complex conjugate gradient $\nabla_{\conj{z}}$, which coincides
with the classical gradient of the same function viewed as $\real^2 \to \real$,
(see appendix~\ref{sub:wirtinger_calculus}).
%
% in natively $\cplx$-valued backprop it would be necessary to modify the existing sgd
% to explicitly use $\nabla_{\conj{z}} f(z)$, while no such modification is necessary
% in re-im. If some inner activation of a $\cplx$VNN is holomorphic, i.e. satisfies
% Cauchy-Riemann conditions, then that particular step of in the chain rule that is
% backprop can be simplified by assuming one piece of the tracked gradient to be exact
% zero. But other wise it is necessary to track both $\partial_{\conj{z}}$ and $\partial_{z}$.
%
% Training a $\cplx$VNN involves optimizing a real-valued objective with respect to
% complex valued arguments.
%
% The computed gradients are theoretically justified
% - CR dictates the it is necessary to track both $z$ and $\conj{z}$ derivatives
% - the direction of largest increase in f: C-R is given by the gradient with
% respect to the conjugate, which coincides in the re-im representation with 
% the ordinary gradient wrt u and v
% - the Chain Rule in CR holds and when a holomorphic C-C function is backproped
% the autodiff just wastes cycles on computing its conj-derivative (rather than assuming
% it to be zero)

% Although $\cplx$-valued dense layers and back-propagation have been around since early
% 1990-s, one of the earliest accounts
Development of deep $\cplx$-valued networks has been active. \citet{haensch_complex-valued_2010}
put forward $\cplx$-valued convolutional networks, \citet{guberman_complex_2016} and
\citet{popa_complex-valued_2017} developed modifications of pooling, \citet{arjovsky_unitary_2016}
and \citet{wisdom_full-capacity_2016} proposed $\cplx$-valued RNNs with unitary recurrent
transition matrices, and \citet{danihelka_associative_2016} developed $\cplx$-valued
holographic representations for LSTMs. More recently \citet{trabelsi_deep_2018} proposed
$\cplx$-valued batch-normalization and weight initialization, \citet{wolter_complex_2018}
investigated different $\cplx$-valued gating mechanisms for RNNs, and \citet{yang_complex_2019}
proposed $\cplx$-valued self-attention and complex transformer architecture. It merits
noting that \citet{gaudet_deep_2018} generalized $\cplx$VNN further to deep quaternion-valued
networks, and \citet{vecchi_compressing_2020} studied sparsity inducing regularizers for
them.
% complex-valued distributions \citep{pav_moments_2015,taubock_complex-valued_2012},
% and \citep{karseras_caution:_2014}

@@

In this section we develop Sparse Variational Dropout for $\cplx$VNN by using a fully
factorized complex Gaussian posterior approximation. We outline the $\cplx$ version of
the local reparameterization trick and derive the divergence penalties in \eqref{eq:sgvb_estimator}.
The proposed $\cplx$-valued extension can readily be a part of a hierarchical variational
approximation for structured sparsity \citep{louizos_bayesian_2017}.

@@

$, respectively, and the random vector $z$ is a \emph{circularly symmetric} $\cplx$-Gaussian
vector if $z$ and $\conj{z}$ are uncorrelated, i.e. $C = 0$.

@@

For a fully factorized approximation $q(\omega)$ and factorized prior $
  \pi(\omega) = \prod_{ij} \pi(\omega_{ij})
$, the divergence term in ELBO \eqref{eq:sgvb_estimator} is
\begin{equation}  \label{eq:elbo-general-kl-div}
  KL(q \| \pi)
    % = \mathbb{E}_{\oemga \sim q(\oemga)} \log \tfrac{q(\oemga)}{\pi(\oemga)}
    % = \sum_{ij} \mathbb{E}_{\omega_{ij} \sim q(\omega_{ij})}
    %   \log \tfrac{q(\omega_{ij})}{\pi(\omega_{ij})}
    % = \sum_{ij} KL\bigl( q(\omega_{ij}) \| \pi(\omega_{ij}) \bigr)
    = - \sum_{ij}
        \mathbb{H}(q(\omega_{ij}))
        + \mathbb{E}_{q(\omega_{ij})} \log{\pi(\omega_{ij})}
    \,.
\end{equation}
We consider two fully factorized priors: an improper prior, resembling VD, and $\cplx$-Gaussian
ARD prior. We omit subscripts ${ij}$ for brevity in this section.

@@

% write a note on the incompressible parameters (bias, batch norm etc.)
The compression rate is calculated based on the number of floating point values needed
to store the network and equals $
  \tfrac{n_\mathtt{par}}{n_\mathtt{par} - n_\mathtt{zer}}
$, where $n_\mathtt{zer}$ is the number of explicit zeros at the ``fine-tune'' stage
and $n_\mathtt{par}$ is the total number of values. In a $\real$-valued network each
parameter counts as \emph{one} value and as \emph{two} values in a $\cplx$VNN. Each
model has a compression limit, determined by biases, shift and scaling in $\real$-
and $\cplx$-valued batch normalization layers.
% variace parameters in q_\theta are completely ignored.
